---
title: "Establishing HLA Typing at CUBI and Benchmarking Tools"
author: 
- "Clemens Messerschmidt"
- "Manuel Holtgrewe"
date: "13.10.2015"
output:
  pdf_document:
    includes:
      in_header: header.tex
bibliography: hla.bib
---

# Abstract
The human leucocyte antigen (HLA) gene cluster plays an important role in adaptive immunity.
HLA class I genes encode for proteins on every nucleated cell that present peptides from inside the cell.

# Introduction

In clinical use, HLA typing often requires the use of special techniques for the enrichment of the HLA gene cluster.
As this requires a commitment of resources and time, it would be desirable to gather this information from already
existing information, e.g. whole exome sequences (WES).

# Tools

- @szolek2014 presented OptiType.
- @bai2014 PHLAT
- Li (unpublished) BWAkit
- @wittig2015 HLAssign

input     | hlassign | optitype | bwakit
----------|----------|----------|--------
HLA panel | \cmark   | \cmark   | 
WES       | ?        | \cmark   | ?
WGS       |          | \cmark (yara) | \cmark

Table: Tool applicability based on type of NGS data

# Datasets

To benchmark the presented tools, multiple dataset were gathered, covering all 
3 types once. For the benchmark, we rely on NGS datasets with known HLA types, 
i.e. the individuals were additionally HLA typed using conventional methods like
Sanger sequencing.

### HLA Panel

@wittig2015 presented HLAssign, a tool to predict HLA types together with a 
custom panel to enrich for the HLA gene cluster. The supplementary material for 
the publication includes sequencing datasets from different sequencing system, 
i.e. Hiseq, Miseq.For our benchmark, we analyzed 311 Hiseq panel datasets, which
were sequenced paired-end, 100 bp.

_CAUTION_: The datasets only contain reads that have a perfect match seed of 30 
against the database. Even though the reads are paired, the fastq files are _not_.
This can introduce problems with aligners like Bowtie2, which is used by PHLAT. 
In this case, reads were treated as single-end. HLAssign, OptiType and bwakit 
identify reads pairs by their name.

_SAMPLE SWAPS_ In a pilot study we were able to discover sample swaps in the data,
i.e. the FASTQ files were mixed up for ~ 10 % of all samples. This leads to a 
decrease in performace, as correctly predicted HLA types will be classified as 
false as they do not match the reference.

### WES

#### Sample set 1: 11 datasets.

The exome-seq datasets are from the 1000 Genomes Project 
(ftp-trace.ncbi.nih.gov/1000genomes/ftp/data/) and were used
by @liu2013 (Supplementary table 2 http://nar.oxfordjournals.org/content/suppl/2013/05/11/gkt481.DC1/nar-00605-met-k-2013-File004_update.pdf).
Samples were prepared using Agilent SureSelect V2 and provide 96-98 % 
exome coverage (>= 10X), run on Illumina Genome Analyzer IIx.

#### Sample set 2: 182 datasets.

First used by @major2013 and later by @szolek2014, this dataset contains 182
exomes, also from the 1000 Genomes Project.

From @major2013:

> There are 270 Coriell IDs in the HapMap database, but at the end of QC check 
> there were only 31 Coriell IDs (41 samples) for whole-genome experiments and 
> 131 Coriell IDs (182 samples) for whole-exome experiments left – many Coriell 
> cell lines were sequenced more than once. Figure 1 explains the details of 
> this filtering process.

### WGS

20 samples of low-coverage WGS data (~ 30X) of the HapMap Project, also 
used by @warren2012, were used for benchmarking on WGS. This data are
2 x 102 bp reads, sequenced on Illumina HiSeq 2000.

# Benchmarks

Performance was evaluated using measures from classification theory with the
problem treated as multi-label, multi-class. Each item, i.e. a sample or 
patient can be assigned with labels signalling one HLA allele each by each 
method. Duplicates are allowed for homozygous predictions.
Lables are than compared to the reference HLA type.

$TP_i$ is number of alleles classified as $type_i$ in reference and prediction

$FP_i$ is number of alleles classified as $type_i$ in the prediction but not in the reference

$FN_i$ is number of alleles classified as $type_i$ in the reference but not in the prediction

The definitions for recall ($\rho$), precision ($\pi$) and F-measure are taken from @ozgur2005:

$\rho_i = \frac{TP_i}{TP_i+FN_i}$

$\pi_i = \frac{TP_i}{TP_i+FP_i}$

The F-measure is a measure for a test's accuracy. It considers both precision 
and recalls, traditionally their harmonic mean (F1-measure) and will take values
between 0 and 1. 
The overall F-measure over all categories (HLA types) can be computed in two fashions.
Here, we opt for the micro-averaged F-measure, which gives equal weight to each
sample. It tends to be dominated by the classifiers (i.e. HLA typing method) 
performance on common categories (e.g. A*02:01) rather than giving equal
weight to categories. The latter would mean that the performance on rare
HLA types would be equally important as on frequent ones.

The micro-averaged F-Measure for $N$ categories is computed as follows:

$\rho = \frac{TP}{TP+FN} = \frac{\sum_{i}^N TP_i}{\sum_{i}^N TP_i+FN_i}$

$\pi =  \frac{TP}{TP+FP} = \frac{\sum_{i}^N TP_i}{\sum_{i}^N TP_i+FP_i}$

$F_1 = \frac{2 \rho \pi}{\rho + \pi}$

# Performance on Public Datasets

The following section will list the performance results for the presented
measures on each of the datasets. If a method is missing from in the table
no results were generated as of the writing of this report. If necessary,
specifics for the results of a given method on a dataset will be reported, if
necessary for the correct interpretation of performance.

# HLA Panel

```{r, echo=FALSE}
library(knitr)
setwd("~/dev/hla/2015-10-01")
source("~/dev/hla/accordance.R")

kable(performance, digits = 2)
```


_REMARK_: For OptiType, only 306 out of 311 panels were typed because of
missing results (unsolved memory issues).

# WES: Sample set 1 (11 datasets)
```{r, echo=FALSE}
library(knitr)
setwd("~/dev/hla/2015-09-16")
source("~/dev/hla/accordance.R")

kable(performance, digits = 2)
```

# WES: Sample set 2 (182 datasets)
```{r, echo=FALSE}
library(knitr)
setwd("~/dev/hla/2015-09-22")
source("~/dev/hla/accordance.R")

kable(performance, digits = 2)
```

# WGS (20 datasets)
```{r, echo=FALSE}
library(knitr)
setwd("~/dev/hla/2015-10-08")
source("~/dev/hla/accordance.R")

kable(performance, digits = 2)
```

# Internal Results

# References
