---
title: 'HLA typing: whole exomes'
author: "Clemens Messerschmidt"
date: "21.09.2015"
output: html_document
---

For 11 exomes, which were used in the ATHLATES paper and have known HLA types, we performed HLA typing with 
the tool optitype.

```{r, echo=F}
source('~/dev/hla/accordance.R')
```

The following table shows the reference types and the optitype predictions with 4-digit resolution:
```{r}
print(all_alleles)
```

For all alleles that possess a precise enough reference, prediction and reference were compared. The number of possible hits is the number of
alleles per sample that is typed with a precision >= 4-digit, e.g. HLA-A*02:01
```{r}
print(accordance)
colSums(accordance)
colSums(accordance)/ colSums(accordance)[ncol(accordance)]
```

On our small sample set of 11 wes, optitype correctly predicts 66/66 for 2-digit and 41/41 for 4-digit resolution, i.e. 100 % accordance with the reference.
HLAssign show a clearly weaker performance with ~80 % recall.

```{r}
print(rbind(accordance, colSums(accordance), (colSums(accordance)/ colSums(accordance)[ncol(accordance)])*100))
```