---
title: "HLA typing: whole exomes"
author: "Clemens Messerschmidt"
date: "21.09.2015"
output: pdf_document
---

For 12 panels, which were used in the HLAssign paper and have known HLA types, we performed HLA typing with 
the tools optitype, bwakit, hlassign.

```{r, echo=F}
source('~/dev/hla/accordance.R')
```

The following table shows the reference types and the optitype predictions with 4-digit resolution:
```{r}
print(all_alleles)
```

For all alleles that possess a precise enough reference, prediction and reference were compared. The number of possible hits is the number of
alleles per sample that is typed with a precision >= 4-digit, e.g. HLA-A*02:01
```{r}
print(accordance)

colSums(accordance)
(colSums(accordance)/ colSums(accordance)[4])*100
```

3 samples can be identified with having wrong reference types assigned to them. To compute performance, we remove them and redo the calculations.

```{r}
accordance = accordance[-c(5,19,12),]
colSums(accordance)
(colSums(accordance)/ colSums(accordance)[4])*100
```

Clearly, optitype and hlassign show better performance than bwakit for our sample set of 20 panels (HLA enriched).